---
title: "Reproducible Research: Peer Assessment 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Basic settings
```{r}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
```

### Data Processing
```{r}
setwd("C:/Users/Ahmad Naimi/Desktop")
stormData <- read.csv("C:/Users/Ahmad Naimi/Desktop/repdata_data_StormData.csv")
dim(stormData)
head(stormData, n = 2)
```

###Summary of Data
```{r}
summary(stormData)
```

###Data Processing (for both questions)
```{r}
# Trim the data set to required columns only
stormEvent <- stormData[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", 
    "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

# Create subset for Question 1 and Question 2

# Select data for Fatalities and injuries for Question 1
eventHealth <- subset(stormEvent, !stormEvent$FATALITIES == 0 & !stormEvent$INJURIES == 
    0, select = c(EVTYPE, FATALITIES, INJURIES))

# Select data for Property Damage and Crop Damage for Question 2
eventEconomic <- subset(stormEvent, !stormEvent$PROPDMG == 0 & !stormEvent$CROPDMG == 
    0, select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
Data Processing

This will prepare requiured data to present most harmful events with respect to population health.

```{r}
# Create separate data set for Injury and Fatalities Fatalities
eventHealth_Death <- aggregate(eventHealth$FATALITIES, by = list(eventHealth$EVTYPE), 
    FUN = sum)
# Give proper name for columns
colnames(eventHealth_Death) <- c("EVENTTYPE", "FATALITIES")

# Injury
eventHealth_Inj <- aggregate(eventHealth$INJURIES, by = list(eventHealth$EVTYPE), 
    FUN = sum)
# Give column name
colnames(eventHealth_Inj) <- c("EVENTTYPE", "INJURIES")

# Let's reorder 2 dataset and filter top 5 events for both dataset
eventHealth_Death <- eventHealth_Death[order(eventHealth_Death$FATALITIES, decreasing = TRUE), 
    ][1:5, ]

eventHealth_Inj <- eventHealth_Inj[order(eventHealth_Inj$INJURIES, decreasing = TRUE), 
    ][1:5, ]
```
###Results
```{r}
# plot top 5 events for fatalities and injuries

# Plot Fatalities and store at Death_plot
Death_plot <- ggplot() + geom_bar(data = eventHealth_Death, aes(x = EVENTTYPE, 
    y = FATALITIES, fill = interaction(FATALITIES, EVENTTYPE)), stat = "identity", 
    show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    xlab("Harmful Events") + ylab("No. of fatailities") + ggtitle("Top 5 weather events causing fatalities") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

# Plot injuries and store at variable Inj_plot
Inj_plot <- ggplot() + geom_bar(data = eventHealth_Inj, aes(x = EVENTTYPE, y = INJURIES, 
    fill = interaction(INJURIES, EVENTTYPE)), stat = "identity", show.legend = F) + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) + xlab("Harmful Events") + 
    ylab("No. of Injuries") + ggtitle("Top 5 weather events causing Injuries") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

# Draw two plots generated above dividing space in two columns

grid.arrange(Death_plot, Inj_plot, ncol = 2)
```

We can easily see that Tornado is the major cause with respect to population health, both for causing fatalities and injuries.
2. Across the United States, which types of events have the greatest economic consequences?
Data Processing

This will prepare requiured data to present most harmful events with respect to the economic damages.

```{r}
# select required entries for economy
eventEconomic <- subset(eventEconomic, eventEconomic$PROPDMGEXP == "K" | eventEconomic$PROPDMGEXP == 
    "k" | eventEconomic$PROPDMGEXP == "M" | eventEconomic$PROPDMGEXP == "m" | 
    eventEconomic$PROPDMGEXP == "B" | eventEconomic$PROPDMGEXP == "b")

eventEconomic <- subset(eventEconomic, eventEconomic$CROPDMGEXP == "K" | eventEconomic$CROPDMGEXP == 
    "k" | eventEconomic$CROPDMGEXP == "M" | eventEconomic$CROPDMGEXP == "m" | 
    eventEconomic$CROPDMGEXP == "B" | eventEconomic$CROPDMGEXP == "b")

# Convert ecnomic values to number
eventEconomic$PROPDMGEXP <- gsub("m", 1e+06, eventEconomic$PROPDMGEXP, ignore.case = TRUE)
eventEconomic$PROPDMGEXP <- gsub("k", 1000, eventEconomic$PROPDMGEXP, ignore.case = TRUE)
eventEconomic$PROPDMGEXP <- gsub("b", 1e+09, eventEconomic$PROPDMGEXP, ignore.case = TRUE)
eventEconomic$PROPDMGEXP <- as.numeric(eventEconomic$PROPDMGEXP)
eventEconomic$CROPDMGEXP <- gsub("m", 1e+06, eventEconomic$CROPDMGEXP, ignore.case = TRUE)
eventEconomic$CROPDMGEXP <- gsub("k", 1000, eventEconomic$CROPDMGEXP, ignore.case = TRUE)
eventEconomic$CROPDMGEXP <- gsub("b", 1e+09, eventEconomic$CROPDMGEXP, ignore.case = TRUE)
eventEconomic$CROPDMGEXP <- as.numeric(eventEconomic$CROPDMGEXP)
eventEconomic$PROPDMGEXP <- as.numeric(eventEconomic$PROPDMGEXP)

# then sum the damages by each event type
eventEconomic$TOTALDMG <- (eventEconomic$CROPDMG * eventEconomic$CROPDMGEXP) + 
    (eventEconomic$PROPDMG * eventEconomic$PROPDMGEXP)

eventEconomic <- aggregate(eventEconomic$TOTALDMG, by = list(eventEconomic$EVTYPE), 
    FUN = sum)

colnames(eventEconomic) <- c("EVTYPE", "TOTALDMG")

# Rank the event type by highest damage cost and take top 5 columns
eventEconomic <- eventEconomic[order(eventEconomic$TOTALDMG, decreasing = TRUE), 
    ]
eventEconomic <- eventEconomic[1:5, ]
```

```{r}
# Now plot the graph
ggplot() + geom_bar(data = eventEconomic, aes(x = EVTYPE, y = TOTALDMG, fill = interaction(TOTALDMG, 
    EVTYPE)), stat = "identity", show.legend = F) + theme(axis.text.x = element_text(angle = 30, 
    hjust = 1)) + xlab("Event Type") + ylab("Total Damage")
```

We can easily see that Flood is the major cause with respect to cost of damage.
Summary

    From these data and grpahs, we found that Tornado are most harmful with respect to population health, while Flood have the greatest economic consequences.

